---
title: "10x scRNA-seq Seurat pipeline for 2iLIF, nEnd and 3D nEnd"
author: "Madeleine Linneberg-Agerholm"
---

## Introduction

QC, normalization, scaling and dimensionality reduction for whole dataset including 2iLIF, nEnd and 3D nEnd (originally named endoid)

## Load libraries

```{r message=FALSE}
library(tidyverse)
library(Seurat)
library(patchwork)

random_seed <- 12345
```

```{r message=FALSE}
data <- readRDS("../data/processed/00_merged.rds")
cc <- readRDS("../data/external/mouse_cell_cycle_genes.rds")
```

## QC

### Before filtering
```{r}
data[["percent.mt"]] <- PercentageFeatureSet(data, pattern = "^mt-")
VlnPlot(data, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
FeatureScatter(data, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(data, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```

```{r}
print(paste0("Before filtering: ", dim(data)[2], " cells ",  dim(data)[1], " genes"))
```

## Remove conditions

```{r}
data <- subset(data, Stage != "C_TSC")
```

```{r}
nbins <- 100
min_genes <- 500
max_genes <- 4500
min_cells <- 500
max_cells <- 18000

ggplot(data@meta.data, aes(x=nCount_RNA)) + 
  geom_histogram(bins = nbins) + 
  geom_vline(aes(xintercept=min_cells), linetype="dashed", color='red') + 
  geom_vline(aes(xintercept=max_cells), linetype="dashed", color='red')

ggplot(data@meta.data, aes(x=nFeature_RNA)) + 
  geom_histogram(bins = nbins) + 
  geom_vline(aes(xintercept=min_genes), linetype="dashed", color='red') +
  geom_vline(aes(xintercept=max_genes), linetype="dashed", color='red')

ggplot(data@meta.data, aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mt)) + 
  geom_point() + 
  scale_color_continuous(type = "viridis") + 
  geom_vline(aes(xintercept=min_cells), linetype="dashed", color='red') +
  geom_vline(aes(xintercept=max_cells), linetype="dashed", color='red') +
  geom_hline(aes(yintercept=min_genes), linetype="dashed", color='red') +
  geom_hline(aes(yintercept=max_genes), linetype="dashed", color='red')
```

## Filter

```{r}
normalized_data <- subset(data, subset = 
                    nFeature_RNA > min_genes & nFeature_RNA < max_genes & 
                    nCount_RNA > min_cells & nCount_RNA < max_cells & 
                    percent.mt < 25)
```

```{r QC_after_filtering}
VlnPlot(normalized_data, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 4)
FeatureScatter(normalized_data, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```

```{r}
print(paste0("After filtering: ", dim(normalized_data)[2], " cells ",  dim(normalized_data)[1], " genes"))
```

## Normalize data

```{r}
normalized_data <- NormalizeData(normalized_data)
normalized_data <- FindVariableFeatures(normalized_data)
```

## Identify the 10 most highly variable genes
```{r}
top10 <- head(VariableFeatures(normalized_data), 10)
```

### Plot variable features with and without labels
```{r}
plot1 <- VariableFeaturePlot(normalized_data)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```
## Scaling the data

```{r}
all.genes <- rownames(normalized_data)
normalized_data <- ScaleData(normalized_data, features = all.genes)
```

## Perform linear dimensional reduction

```{r}
normalized_data <- RunPCA(normalized_data, features = VariableFeatures(object = normalized_data))
```

```{r}
ElbowPlot(normalized_data, ndims = 30)
```

```{r}
normalized_data <- CellCycleScoring(normalized_data, s.features = cc$s.genes, g2m.features = cc$g2m.genes, set.ident = FALSE)
```

```{r}
DimPlot(normalized_data, group.by = "Stage")
DimPlot(normalized_data, group.by = "Phase")
```

```{r}
normalized_data <- FindNeighbors(normalized_data, dims = 1:20)
normalized_data <- FindClusters(normalized_data, random.seed = 12345, resolution = 0.6) #0.2 is too low and 0.8 too high, use 0.6
normalized_data <- RunUMAP(normalized_data, reduction = "pca", dims = 1:20, seed.use = 12345)
```

## Plot UMAP

```{r}
DimPlot(normalized_data, reduction = "umap", group.by = "Stage")
DimPlot(normalized_data, reduction = "umap", group.by = "Phase")
DimPlot(normalized_data, reduction = "umap")
```

## Save session

```{r}
saveRDS(normalized_data, file = "../data/processed/normalized_data.rds")
```

## Session info

```{r session-info, cache = FALSE}
devtools::session_info()
```
